# Analysis of longitudinal data

### The RATS dataset

We begin by reading in the RATS data set:

```{r}
RATSL <- read.csv("data/RATSL.csv", row.names = 1)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

dim(RATSL)
str(RATSL)
```

The RATS data set consists of 176 observations of 5 variables (of longitudinal data). The data is from a study on the weight gain of three groups of rats (16 rats total) on different diets over a period of 64 days.

We plot the weight developments of the rat groups as follows:

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line() +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

From the plots we see that the rats in Group 1 are about half the weight of the 
rats in the other groups, and there are twice as many of them. In Group 2, we see that there is an outlier rat of a much heavier weight than the others. Overall, it appears that all rats gained weight during the study, but rats in Groups 2 and 3 gained more absolute weight than the rats in Group 1.

To compare the weight gain better, let us standardize the weights: 

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID)) +
  geom_line() +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")
```

The standardized weight plots tell us that the rats in Group 2 appeared to gain relatively more weight on average than rats in the other groups. Compared to their initial weights, rats in Group 3 appear to have gained the least weight, with the rats in Group 2 positioned somewhere in between.

Next, we will look at the mean response profile.

```{r}
n <- RATSL$Time %>% unique() %>% length()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

From the mean response profile plot we can tell that the groups are clearly different from one another, because the standard errors do not overlap. We can also see what we already noted on earlier: the mean weights of each group increase during the study, and the rats in Group 2 appear to be gaining weight faster than the rats in Group 3, since the corresponding lines in the plot are getting closer to each other with time.

Let us then consider the mean summary measure. 

```{r}
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean = mean(Weight)) %>%
  ungroup()

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 2-9")
```

There appears to be more variability in Group 2, which is understandable due to the heavy outlier we noted on earlier, seen here as a black dot at a weight of almost 600 grams. We also see some less drastic outliers in the other groups. Since these outliers may bias our analysis, we'll filter them out:

```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean > 250 & mean < 550 & (mean < 475 | mean > 525))

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 2-9")
```

We have no more outliers. We can see some more variability with Groups 1 and 2 when compared to Group 3, but we'd like to perform a formal test to be sure. Let us do exactly that. We'll consider the observations of day 1 as a baseline in what follows.

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep="", header=TRUE)

RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)

anova(fit)
```

By looking at the p-values, we can see that the initial weights, i.e. the baseline, has a strong correlation with the mean summary measure. This makes sense. But we can also see that the Group variable has a p-value of 0.07586, which implies that the choice of group actually has no real statistical significance on the mean.

### The BPRS dataset

Let us consider the BPRS data set next. We read in the data:

```{r}
BPRSL <- read.csv("data/BPRSL.csv", row.names = 1)
BPRSL$subject <- BPRSL$subject + 20*(BPRSL$treatment == 2)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

dim(BPRSL)
str(BPRSL)
```

The BPRS data set consists of 176 observations of 6 variables (of longitudinal data). The data records the brief psychiatric rating scale (bprs) measured from 40 male subjects divided into two treatment groups, over a period of 8 weeks.

We plot the psychiatric rating over the treatment period as follows:

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

It looks like in the case of treatment 1, the bprs rating goes down during the treatment for most subjects. Things are a little less clear for the treatment 2 group, and we see that we again have an outlier in the second group.

We will ignore the longitudinal nature of the data for now. We fit a multiple linear regression model, where bprs is the target variable and week + treatment are the explanatory variables. We find the following model:

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

Our naive first analysis tells us that time has a strong impact on the bprs rating, but the choice between the treatment groups has no statistical significance. 

To make our model more realistic, we must take into account the fact that the repeated measures are not independent. We will thus fit a random intercept model for the same two explanatory variables, and then a random intercept and random slope model:

```{r}
library(lme4)

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)

BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)

anova(BPRS_ref1, BPRS_ref)
```

From the chi-squared statistics and the p-value we see that the second model, the random intercept and random slope model, is a better fit to the data. However, we can still go a bit further in our analysis: let us take into account the group x time interaction and fit a random intercept and random slope model:

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)

Fitted <- fitted(BPRS_ref2)

BPRSL <- BPRSL %>%
  mutate(Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$Fitted), max(BPRSL$Fitted)))
```

Interestingly, from the chi-squared statistics and the p-value we can see that this final model is actually a worse fit than our previous one. Our interpretation is that there is no real group x time interaction in the data.  