# Analysis of longitudinal data

### The RATS dataset

We begin by reading in the RATS data set:

```{r}
RATSL <- read.csv("data/RATSL.csv", row.names = 1)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

dim(RATSL)
str(RATSL)
```
The RATS data set consists of 176 observations of 5 variables (of longitudinal data). The data is from a study on the weight gain of three groups of rats (16 rats total) on different diets over a period of 64 days.

We plot the weight developments of the rat groups as follows:

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line() +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

From the plots we see that the rats in Group 1 are about half the weight of the 
rats in the other groups, and there are twice as many of them. In Group 2, we see that there is an outlier rat of a much heavier weight than the others. Overall, it appears that all rats gained weight during the study, but rats in Groups 2 and 3 gained more absolute weight than the rats in Group 1.

To compare the weight gain better, let us standardize the weights: 

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID)) +
  geom_line() +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")
```

The standardized weight plots tell us that the rats in Group 2 appeared to gain relatively more weight on average than rats in the other groups. Compared to their initial weights, rats in Group 3 appear to have gained the least weight, with the rats in Group 2 positioned somewhere in between.

Next, we will look at the mean response profile.

```{r}
n <- RATSL$Time %>% unique() %>% length()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

From the mean response profile plot we can tell that the groups are clearly different from one another, because the standard errors do not overlap. We can also see what we already noted on earlier: the mean weights of each group increase during the study, and the rats in Group 2 appear to be gaining weight faster than the rats in Group 3, since the corresponding lines in the plot are getting closer to each other with time.

Let us then consider the mean summary measure. 

```{r}
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean = mean(Weight)) %>%
  ungroup()

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 2-9")
```

There appears to be more variability in Group 2, which is understandable due to the heavy outlier we noted on earlier, seen here as a black dot at a weight of almost 600 grams. We also see some less drastic outliers in the other groups. Since these outliers may bias our analysis, we'll filter them out:

```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean > 250 & mean < 550 & (mean < 475 | mean > 525))

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 2-9")
```

We have no more outliers. We can see some more variability with Groups 1 and 2 when compared to Group 3, but we'd like to perform a formal test to be sure. Let us do exactly that. We'll consider the observations of day 1 as a baseline in what follows.

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep="", header=TRUE)

RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)

anova(fit)
```

### The BPRS dataset

We read in the BPRS data set:

```{r}
BPRSL <- read.csv("data/BPRSL.csv", row.names = 1)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Linear regression model:

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

Random intercept model:

```{r}
library(lme4)

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

Random intercept and random slope model:

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

Random intercept and random slope model with interaction:

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

Fitted <- fitted(BPRS_ref2)

BPRSL <- BPRSL %>%
  mutate(Fitted)

#ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
#  geom_line(aes(linetype = treatment)) +
#  scale_x_continuous(name = "week", breaks = seq(0, 60, 20)) +
#  scale_y_continuous(name = "Fitted bprs") +
#  theme(legend.position = "top")

ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$Fitted), max(BPRSL$Fitted)))
```

